Supporting Information:\\ NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2021-04-06 manuscriptPGPEsuppl.tex
Updates from Overleaf

\author{\scriptsize Am{\'e}lie Bacle}
\affiliation{Paris, France}

@@

\author{Patrick Fuchs}
\affiliation{Paris, France}

@@

\author{Jesper J. Madsen}
\affiliation[University of Chicago]{Department of Chemistry, The University of Chicago, Chicago, Illinois, United States of America}
\alsoaffiliation[University of South Florida]{Global and Planetary Health, College of Public Health, University of South Florida, Tampa, Florida, United States of America}

@@

\author{Paula Milan Rodriguez}
\affiliation{Paris, France}

@@

\author{Chris G. Papadopoulos}
\affiliation[]{I2BC - University Paris Sud}

@@

Headgroup order parameters of PC lipids are unchanged upon addition
of zwitterionic lipids or cholesterol in experiments, but increase
upon addition of negatively charged PG or PS lipids because
headgroup dipole tilts more parallel to the membrane plane after incorporation
of negative charges into the membrane~\cite{seelig87, scherer87,antila18}.
The response of PC headgroup order parameters to the addition of PE or PG lipids
from different simulations is compared with experiments in figure~\ref{HGorderparametersPCvsPEPG}.
None of the simulations reproduce neither the experimentally observed increase in PC headgroup order parameters
with increasing amount of PG nor the related tilting of the headgroup more parallel with the membrane.
Similar observations in our previous work for PS lipids were explained by the overestimated counterion
binding affinity that neutralizes the effect of added negative charge \cite{antila19}.
All simulations except Berger-OPLS predict tilting of P-N headgroup outwards from the membrane and
decrease of PC headgroup order parameters upon addition of PE lipids.
These results are not in line with experiments where the PC headgroup order parameters are not affected by zwitterionic lipids \cite{scherer87}.
The good performance of Berger-OPLS simulations is surprising here because
headgroup conformational enemble is not very close to experiments in this model and
the response of headgroup order parameters
to cholesterol was significantly overestimated by the Berger/H{\"o}ltje force field in our previous work \cite{botan15}.

@@

\noindent {\it POPC:POPE mixtures} Data is available at \cite{POPCcharmm300K,POPC1POPE1charmm36}.
300 K with v-rescale (tau=0.1 ps),
1 bar with PR semiisotropic (tau=4 ps, compressibility=4.5e-5 bar$^{-1}$),
PME order 4 and space 0.12,
rcoulomb and rvdw 1.0,
128 lipids per leaflet,
no ion \todo{Full simulation details by Fuchs et al.}

@@

\noindent {\it POPC:POPE mixtures} Data is available at \cite{POPCberger300K,POPC1POPE1berger}. 
300 K with v-rescale (tau=0.1 ps),
1 bar with PR semiisotropic (tau=4 ps, compressibility=4.5e-5 bar$^{-1}$),
PME order 4 and space 0.12,
rcoulomb and rvdw 1.0,
128 lipids per leaflet,
no ion \todo{Simulation details by Fuchs et al.}

@@

\noindent {\it POPC:POPE mixtures} \todo{Simulation details by P. Fuchs.}

@@

\section{Author contributions}
\noindent
{\it Am{\'e}lie Bacle}
%% Please write description of your contribution here

@@

\noindent
{\it Patrick Fuchs}
%% Please write description of your contribution here

@@

\noindent
{\it Jesper J. Madsen}
set up, performed, and analysed several of the CHARMM36 simulations.
Provided comments on the manuscript.

@@

\noindent
{\it Paula Milan Rodriguez}
%% Please write description of your contribution here

@@

\noindent
{\it Chris G. Papadopoulos}
%% Please write description of your contribution here

>>>>

\author{\scriptsize Am{\'e}lie Bacle}
\affiliation{Laboratoire Coopératif "Lipotoxicity and Channelopathies – ConicMeds", Université de Poitiers, 1 rue Georges Bonnet, 86000 Poitiers, France }

@@

\author{Patrick F.J. Fuchs}
\affiliation{Sorbonne Université, Ecole Normale Supérieure, PSL University, CNRS, Laboratoire des Biomolécules (LBM), 75005 Paris, France}
\alsoaffiliation{Université de Paris, UFR Sciences du Vivant, 75013, Paris, France}

@@

\author{Jesper J. Madsen}
\affiliation[University of Chicago]{Department of Chemistry, The University of Chicago, Chicago, Illinois, United States of America}
\alsoaffiliation[University of South Florida]{Department of Global Health, College of Public Health, University of South Florida, Tampa, Florida, United States of America}

@@

\author{Paula Milán Rodríguez}
\affiliation{Sorbonne Université, Ecole Normale Supérieure, PSL University, CNRS, Laboratoire des Biomolécules (LBM), 75005 Paris, France}

@@

\author{Chris G. Papadopoulos}
\affiliation{Université Paris-Saclay, CEA, CNRS, Institute for Integrative Biology of the Cell (I2BC), 91198, Gif-sur-Yvette, France}

@@

Headgroup order parameters of PC lipids are unchanged upon addition
of zwitterionic lipids or cholesterol in experiments, but increase
upon addition of negatively charged PG or PS lipids because
headgroup dipole tilts more parallel to the membrane plane after incorporation
of negative charges into the membrane~\cite{seelig87, scherer87,antila18}.
The response of PC headgroup order parameters to the addition of PE or PG lipids
from different simulations is compared with experiments in figure~\ref{HGorderparametersPCvsPEPG}.
None of the simulations reproduce neither the experimentally observed increase in PC headgroup order parameters
with increasing amount of PG nor the related tilting of the headgroup more parallel with the membrane.
Similar observations in our previous work for PS lipids were explained by the overestimated counterion
binding affinity that neutralizes the effect of added negative charge \cite{antila19}.
All simulations except Berger-OPLS predict tilting of P-N headgroup outwards from the membrane and
decrease of PC headgroup order parameters upon addition of PE lipids.
These results are not in line with experiments where the PC headgroup order parameters are not affected by zwitterionic lipids \cite{scherer87}.
The good performance of Berger-OPLS simulations is surprising here because
headgroup conformational ensemble is not very close to experiments in this model and
the response of headgroup order parameters
to cholesterol was significantly overestimated by the Berger/H{\"o}ltje force field in our previous work \cite{botan15}.

@@

\noindent {\it POPC:POPE mixtures} Data is available at \cite{POPCcharmm300K,POPC1POPE1charmm36}.
A pure POPC system and a 50:50 POPC:POPE mixture were built and equilibrated using CHARMM-GUI [https://doi.org/10.1021/acs.jctc.5b00935]. They contained 256 lipids (for the mixture 64 lipids per leaflet for each species) and 34 water molecules per lipid. No ions were added. The production simulations were run for 300 ns with a time step of 2 fs. The first 50 ns were discarded for the analysis. The simulations were run with the GROMACS 2016.4 [https://doi.org/10.1016/j.softx.2015.06.001] version. The v-rescale thermostat [http://dx.doi.org/10.1063/1.2408420] was used with a temperature of 300 K and a time constant of 1 ps; lipids and water were coupled separately to the heat bath. Pressure was kept constant at 1 bar using a semi–isotropic Parrinello–Rahman barostat [https://doi.org/10.1103/PhysRevLett.45.1196] with a time constant of 5.0 ps. A real space cut-off of 1.2 nm was employed for electrostatic interactions while the long-range part was evaluated using the PME method [https://doi.org/10.1063/1.464397, https://doi.org/10.1063/1.470117]. A force-based switching function was used to switch the Lennard-Jones forces to zero over a range of 1−1.2 nm. All bonds with hydrogen atoms were constrained with the LINCS algorithm [https://doi.org/10.1002/(SICI)1096-987X(199709)18:12%3C1463::AID-JCC4%3E3.0.CO;2-H]. Water molecules were kept rigid with the SETTLE algorithm [https://doi.org/10.1002/jcc.540130805]. The simulation files are available from ref 64 [https://doi.org/10.5281/zenodo.1306800] (pure POPC) and 65 [https://doi.org/10.5281/zenodo.1306821] (POPC:POPE mixture).

@@

\noindent {\it POPC:POPE mixtures} Data is available at \cite{POPCberger300K,POPC1POPE1berger}. 
Two systems were simulated using the Berger force field [https://doi.org/10.1016/S0006-3495(97)78845-3], pure POPC and a mixture 50:50 POPC:POPE. For POPE, we additionnally used the de Vries modification implementing a repulsion of the hydrogen atoms located on the amino group of ethanolamine [https://doi.org/10.1021/jp0366926].
Starting from a PDB file of a pure POPC system with Berger atom names, a 50:50 POPC:POPE mixture was built by mutating randomly methyl groups to hydrogens (POPC -> POPE). Each system contained 256 lipids (for the mixture 64 lipids per leaflet for each species) and about 40 water molecules per lipid. No ions were added. Both systems were minimized and equilibrated. The production simulations were run for 300 ns with a time step of 2 fs. The first 100 ns were discarded for the analysis (@Samuli: this sentence may be removed if the analysis was done on a different time window). The simulations were run with GROMACS 4.5.3 [https://doi.org/10.1093/bioinformatics/btt055] version. The v-rescale thermostat [http://dx.doi.org/10.1063/1.2408420] was used with a temperature of 300 K and a time constant of 0.1 ps; lipids and water were coupled separately to the heat bath. Pressure was kept constant at 1 bar using a semi–isotropic Parrinello–Rahman barostat [https://doi.org/10.1103/PhysRevLett.45.1196] with a time constant of 4.0 ps. A real space cut-off of 1.0 nm was employed for van der Waals and electrostatic interactions. The long-range part of electrostatic interactions was evaluated using the PME method [https://doi.org/10.1063/1.464397, https://doi.org/10.1063/1.470117] with a grid spacing of 0.12 nm and an interpolation order of 4. All bonds with hydrogen atoms were constrained with the LINCS algorithm [https://doi.org/10.1002/(SICI)1096-987X(199709)18:12%3C1463::AID-JCC4%3E3.0.CO;2-H]. Water molecules were kept rigid with the SETTLE algorithm [https://doi.org/10.1002/jcc.540130805]. The simulation files are available from ref 91 [https://doi.org/10.5281/zenodo.1402417] (pure POPC) and 92 [https://doi.org/10.5281/zenodo.1402449] (POPC:POPE mixture).

@@

\noindent {\it POPC:POPE mixtures}

Force field details
Simulations of POPC:POPE mixtures using the OPLS-MacRog force field [https://doi.org/10.1021/jp5016627] were performed. The initial force field parameter files in GROMACS format were taken from ref [https://doi.org/10.1016/j.dib.2016.03.067]. However, a number of errors were detected in the published files so we fixed them in the following way. For POPE, the two aliphatic chains sn-1 and sn-2 were switched (in the file the lipid was in fact OPPE); one atom (named C27) was not connected to the previous atom in the aliphatic chain so bonds, angles, dihedrals, pairs were included to create the connection; two atoms were called C27 thus one of them was renamed C28. For POPC, some impropers were missing leading to non planar systems for double bonds (in particular the carbonyls of sn-1 and sn-2 as well as the double bond of the oleoyl chain), we thus added back those impropers. The resulting fixed files can be found on Zenodo (https://zenodo.org/record/3741793/files/popc_fixed.itp, https://zenodo.org/record/3725637/files/pope_fixed.itp). To check our fix, one simulation of pure POPC was also run and the order parameter compared to previous published results of ref [https://doi.org/10.1021/acs.jpcb.5b04878]. We found very similar values.
All the details of the procedure described here are described in https://www.dsimb.inserm.fr/~fuchs/project_Samuli/POPC_POPE/report_results_comparison.pdf.
The data for pure POPC are available from [https://zenodo.org/record/3741793], for the POPC:POPE mixture from [https://zenodo.org/record/3725637].

Simulation details
Two systems of pure POPC and 50:50 POPC:POPE mixture were built using CHARMM-GUI [https://doi.org/10.1021/acs.jctc.5b00935]. They contained 128 lipids and 5120 TIP3 water molecules. The initial PDB file was modified to match OPLS-MacRog nomenclature and atom order. No ions were added. The production simulations were run for 500 ns with a time step of 2 fs. The first 200 ns were discarded for the analysis (@Samuli: this sentence may be removed if the analysis was done on a different time window). The simulations were run withthe GROMACS 2018.5 [https://doi.org/10.1016/j.softx.2015.06.001] version. The v-rescale thermostat [http://dx.doi.org/10.1063/1.2408420] was used with a temperature of 300 K and a time constant of 0.1 ps; lipids and water were coupled separately to the heat bath. Pressure was kept constant at 1 bar using a semi–isotropic Parrinello–Rahman barostat [https://doi.org/10.1103/PhysRevLett.45.1196] with a time constant of 4.0 ps. Long-range electrostatic interactions were calculated using the PME method [https://doi.org/10.1063/1.464397, https://doi.org/10.1063/1.470117]. A real space cut-off of 1.0 nm was employed with a grid spacing of 0.1 nm in the reciprocal space. Lennard-Jones potentials were cut off at 1.0 nm. All covalent bonds in lipids were constrained using the LINCS algorithm [https://doi.org/10.1002/(SICI)1096-987X(199709)18:12%3C1463::AID-JCC4%3E3.0.CO;2-H]. Water molecules were kept rigid with the SETTLE algorithm [https://doi.org/10.1002/jcc.540130805]. The simulation files are available from ref 66 [https://doi.org/10.5281/zenodo.3741793] (pure POPC) and 67 [https://doi.org/10.5281/zenodo.3725637] (POPC:POPE mixture).

@@

\section{Author contributions}
\noindent
{\it Am{\'e}lie Bacle} Amélie Bacle set up, performed and analysed POPC and POPC:POPE (1:1) simulations with the Berger Force Field.
%% Please write description of your contribution here

@@

\noindent
{\it Patrick F.J. Fuchs} supervised Paula Milán Rodríguez, Amélie Bacle and Chris Papadopoulos, created the buildH software, contributed to many discussions.
%% Please write description of your contribution here

@@

\noindent
{\it Jesper J. Madsen}
%% Please write description of your contribution here

@@

\noindent
{\it Paula Milán Rodríguez} set up, performed and analysed POPC and POPC:POPE (1:1) simulations with the MacRog Force Field. She also fixed the initial force field (itp) files prior to the simulations.
%% Please write description of your contribution here

@@

\noindent
{\it Chris G. Papadopoulos} set up, performed and analysed POPC and POPC:POPE (1:1) simulations with the CHARMM36 Force Field.
%% Please write description of your contribution here
